Investigating the alternative prey hypothesis with the POMP framework

### 2024-04-22 workspace/heyuan.Rmd
Add files via upload

```{r}
library(tidyverse)
library(tseries)
#install.packages("ggfortify")
library(ggfortify)
```

@@

## POMP 1

```{r}
install.packages("pomp")
library(pomp)
```

To create a POMP for CPUE, I need to define the state process that reflects the underlying dynamics of the species population and the observation process that relates the state to the CPUE measurements.

>>>>

```{r}
library(tidyverse)
library(tseries)
#install.packages("ggfortify")
library(ggfortify)
library(pomp)
```

@@

## POMP 3

Y(t): bird count, $y_t = NB(mean = \rho B_t, \sigma)$, $B_t$ is bird population size at time t
X(t): unobserved state process, red foxes, Bird, Peak Rodent Year, temperature, precipitation, NAO index

Reference: Lotka-Volterra predator-prey model for the state process: $$X_{t+dt} = \beta X_t - \mu X_t$$

natural born - number of birds eaten by foxes: 
$$B_{t+dt} = \alpha B_t - \beta(R_t) B_{t}F_{t}$$

where, $$\beta(R_t) = \beta if R(t)=0$$ or 
$$ \beta(R_t) = \gamma \beta, \gamma > 1$$ , rodent number affect the rate of fox eating birds, 
$\beta(R_t)$ can be re-written as $$\beta(R_t) =  \gamma^R \beta$$, where R = 0 or 1

Growth Rate - Death Rate of Foxes: 
$$ F_{t+dt} = \delta(t) F_t B^*_t - \theta(t) F_t$$

where, $$\delta(t) = \delta_1 if R(t) = 0, or, \phi \delta_1$$ , $\phi<1$ and whether it is in rodent peak year affect the number of birds and the number of foxes.

$\delta(t)$ can be re-written as $$\delta(t) = \phi^R \delta_1$$, where R = 0 or 1

$\theta(t)$ could include the temperature, NAO about how does it affect the death rate of Foxes.

first, let dt=1 year
then, dt = less than 1 year, maybe 1 month, bird keeps as 100, foxes=10, 15, 17, 15 for each dt until reaching 1 year. 
then include bird in X(t), lecture 16

another way: bird is not in X(t)

$$B_{t+dt} = \alpha B^*_{t+dt} - \beta(R_t) B_{t} F_{t}$$

just do the fox equation, no bird equation

```{r}
# compute the average temperature and preciptation. 
data_frame <- df[,c(1,3,6,7,8:17)]
data_frame$Temp <- rowMeans(data_frame[,c(5:10)], na.rm = TRUE) 
data_frame$Precip <- rowMeans(data_frame[,c(11:14)], na.rm = TRUE) 
data_frame <- data_frame[,c(1,2,3,4,15,16)]
names(data_frame) <- c('year', 'CUPE', 'R', 'NAO', 'Temp', 'Precip')
data_frame$R <- ifelse(data_frame$R == "yes", 1, 0)
data_frame <- transform(data_frame, year = as.numeric(year))  # Ensure 'year' is numeric
data_frame <- data_frame[c(3:142), ]

Euler Approximation of the predator population at time $t+ \delta t$...

$$F_{t+\delta t} = F_t + \delta(t) F_t B_t \delta t - \theta F_t \delta t$$
$$F_{t+\delta t} = F_t + (\phi^R \delta_1 F_t B_t - \theta F_t) \delta t$$

$\delta(t) F_t B_t \delta t$ is the expected number of new predators arising from the Poisson process with rate $\delta(t) F_t B_t$;
$- \theta F_t \delta t$ is the deterministic decrease in predators due to natural death over the time step $\delta t$

$$B_{t+ \delta t} = B_t + \alpha B_t \delta t - \beta(R_t) B_t F_t\delta t$$
$$B_{t+ \delta t} = B_t + (\alpha B_t - \gamma^R \beta B_t F_t) \delta t$$

```{r}
meas <- data_frame %>% select(year, reports=CUPE, R) %>% as.data.frame()

sir_step <- function (Fox, B, psi, gamma, alpha, delta_1, beta_1, delta.t,theta, ...) {
  dF = rpois(1, pow(psi, R) * delta_1 * Fox * B) 
  dB = rpois(1, alpha * B) - rpois(1, pow(gamma, R) * beta_1 * B * Fox)

  Fox <- Fox + dF - theta * Fox * delta.t
  B <- B + dB
  c(Fox = Fox, B = B)
}

sir_rinit <- function (...) {
  c(Fox = 90, B = 90)
}

meas |>
  pomp(times="year",t0=meas$year[1]-1,
    rprocess=euler(sir_step,delta.t=1),
    rinit=sir_rinit
  ) -> measSIR

# 
sir_step<- Csnippet("
  double dF = rpois(1, pow(psi, R) * delta_1 * Fox * B) 
  double dB = rpois(1, alpha * B) - rpois(1, pow(gamma, R) * beta_1 * B * Fox)

  Fox += dF -theta * Fox * delta.t;  
  B += dB;  
")

# dmeasure: Define the measurement model
sir_dmeas <- Csnippet("
  lik = dnbinom_mu(reports, size = rho * B, prob = exp(-sigma), give_log);
")

# rmeasure: 
sir_rmeas <- Csnippet("
  reports = rnbinom_mu(rho * B, prob = exp(-sigma));
")
## size =  number of successes until the experiment is stopped; success rate / hunting rate:

# initialize:
sir_rinit <- Csnippet("
  Fox = 90;
  B = 90;
")

# Define parameters and initial states
params <- c(alpha = 0.05, beta_1 = 0.02, gamma = 2, delta_1 = 0.01, psi = 3, theta = 0.1, rho = 0.5, sigma = 1)

# create a pomp object
measSIR |>
  pomp(rprocess=euler(sir_step,delta.t=1),
    rinit=sir_rinit,
    rmeasure=sir_rmeas,
    dmeasure=sir_dmeas,
    statenames=c("Fox","B"),
    paramnames=c("alpha", "gamma", "delta_1", 'beta_1', "psi", "theta", "rho" ,'sigma'),
    obsnames = c('R')
  ) -> measSIR

```{r}
meas <- data_frame %>% select(year, reports=CUPE, R) %>% as.data.frame()

sir_step <- function (Fox, psi, delta_1, delta.t, R, B, theta, ...) {
  dF = rpois(1, pow(psi, R) * delta_1 * Fox * B) 
  Fox <- Fox + dF - theta * Fox * delta.t
  Fox = Fox

}

sir_rinit <- function (...) {
  Fox = 90
}

meas |>
  pomp(times="year",t0=meas$year[1]-1,
    rprocess=euler(sir_step,delta.t=1),
    rinit=sir_rinit
  ) -> measSIR

# 
sir_step<- Csnippet("
  double dF = rpois(1, pow(psi, R) * delta_1 * Fox * B);
  Fox += Fox + dF - theta * Fox * delta.t;
")

# dmeasure: Define the measurement model
sir_dmeas <- Csnippet("
  lik = dnbinom_mu(reports, size = rho * B, prob = exp(-sigma), give_log);
")

# rmeasure: 
sir_rmeas <- Csnippet("
  reports = rnbinom_mu(rho * B, prob = exp(-sigma));
")
## size =  number of successes until the experiment is stopped; success rate / hunting rate:

# initialize: number of red foxes at 1983
sir_rinit <- Csnippet("
  Fox = 90;
")

# Define parameters and initial states
params <- c(delta_1 = 0.01, psi = 3, theta = 0.1, rho = 0.5, sigma = 1)

# create a pomp object
measSIR |>
  pomp(rprocess=euler(sir_step,delta.t=1),
    rinit=sir_rinit,
    rmeasure=sir_rmeas,
    dmeasure=sir_dmeas,
    statenames=c("Fox",'B'),
    paramnames=c("delta_1", "psi", "theta", "rho" ,'sigma'),
    obsnames = c('R')
  ) -> measSIR
